Material Metohds

### 2019-09-18 doc/MatMeth.Rmd
added my citation, more cleaning up

To access natural genetic variation for *Mus musculus*, Wild derived inbred strains strains were used. PWD/PhJ, PERC/Eij, WSB/EiJ, LEWE/EiJ, MSM/MsJ, MOLF/EiJ, CAST/EiJ, CZECHII/EiJ, CORLI/EiJ and SPRET/EiJ were purchased from Jax labs (https://www.jax.org) (Maine USA). The strains of KAZ/TUA, TOM/TUA, AST/TUA, HMI/TUA and SPIC) were cryo-derived from Biological Resource Center (BRC) at RIKEN (Ibaraki, Japan) (https://en.brc.riken.jp/). All mice were housed UW-Madison Biotech and MSC facilities - following the protocols.  A breeding colony of wild derived mice sampled from Gough Island (GI )is maintained at UW Veterinary school facilities.
 Mice were feed on dry standard breeder chow. Some strains sunflower seeds, nestlets and larger cages were added to improve fertility and litter survival. Adult mice were euthanized by CO asphyxiation. Neonate and embryonic mice were euthanized by decapitation following the protocols approved by the Institutional Animal Care and Use Committee at the University of Wisconsin-Madison.

Over the coruse of data collection some breeding colonies / mice were moved from facilities. Additionally the GI strain was kept at a seperate facility.  (eviddence for environmental (room) effect on (MLH1 / SC lengths / a variety of meiotic traits was tested for. We observed no effect, Supplemental figure))

@@

From males, the right teste was collected and processed as discribed in X citation. The majority of mice used were betwee 5 and 12 weeks, (supplemental table). However some strains had problems breeding, so testes from older mice were collected. We found a (small/no) age effect on MLH1 counts for male mice (Supplemental Figure)).

The vast majority of oocyte data was collected from neonate litters/mice between 5 to 48 hours. This approach maintain breeding pairing and still result in prophase oocytes (cite timeline of oocytes in neonates). (differences were assessed between embryo samples and neonates?). Presice staging of embryos by checking copulary plugs was diffcult/not feasible in many of these wild derived strains due to their behavior. Embryonic samples were collected when pregenacgy noted in females, embryos were staged based on (X table markers).

Meiocyte spreads were made following [@peters_1997]. For testes the tunica was removed and whole testes was incubated in 3ml of hypotonic solution for 45min. For ovaries after dissection, the pair of ovaries were decapsulated in cold PBS and both were incubated in 300ul of hypotonic solution for 45 min.

After incubation, gonads were transferred to sucrose solution -for mastication to make a cell slurry which was transferred to 80ul of 2% PFA solution. Cells were fixed in this solution and dried in a humid chamber at room temperature overnight. The next morning, slides were treated with a photoflow (Kodak, cite) wash.

@@

The same staining protocol was applied to spermatocyte and oocyte spreads.
The staining protocol was bas on X with some adaptations, and same as previously discribed in Peterson etal 2019.

@@

Hand measures dataset generated by using ImageJ.

@@

The distributions of CO counts per cell was assessed for normailty (supplemental figure). Mouse mean MLH1 count was used to get around the bad discreet nature of count data. (normality was confirmed with X distrbution, supplememtal figure).

-specific statistical tests; (means, non-parameteric kurts wallis t-test?) (correlation coeffectent (function))
linear models lm() from X package.

-specific R packages
-non-parameteric  (tests and measures for dealing with count data)
-hypothese tested (size, DSB number)

Comparison/analysis of variation in MLH1 counts within and between groups was done within a general linear/ mixmodel frame work, using X package (cite).
With X formula Co count ~ blah + blah, with random strain specific sex effect.

(we used this frame work as proxy for 'polymorphism' and 'divergence' in MLH1 counts across house mouse and closeley related species.)

>>>>

To access natural genetic variation for *Mus musculus*, wild derived inbred strains were used. The Mus musculus PWD/PhJ, PERC/Eij, WSB/EiJ, LEWE/EiJ, MSM/MsJ, MOLF/EiJ, CAST/EiJ, CZECHII/EiJ were purchased from Jackson labs (Maine, USA). The strains of KAZ/TUA, TOM/TUA, AST/TUA, and HMI/TUA were cryo-derived from Biological Resource Center (BRC) at Riken (Ibaraki, Japan).  (https://en.brc.riken.jp/).
The related species of *Mus caroli* CAROLI/EiJ and *Mus spretus* SPRET/EiJ were purchased from Jax labs (https://www.jax.org) and *Mus spicilegus* SPIC/Eij from Riken.

 All mice were housed UW-Madison Biotech and MSC facilities - following the protocols.  A breeding colony of wild derived 
*Mus musculus* mice sampled from Gough Island (GI)is maintained at UW Veterinary school facilities.
 Mice were feed on dry standard breeder chow. Some strains sunflower seeds, nestlets and larger cages were added to improve fertility and litter survival. Adult mice were euthanized by CO asphyxiation. Neonate and embryonic mice were euthanized by decapitation following the protocols approved by the Institutional Animal Care and Use Committee at the University of Wisconsin-Madison.

Over the course of data collection some breeding colonies / mice were moved from facilities. Additionally the GI strain was kept at a separate facility.  (evidence for environmental (room) effect on (MLH1 / SC lengths / a variety of meiotic traits was tested for. We observed no effect, Supplemental figure))

@@

Spermatocyte spreads were collected and perpared as described in [@peterson_2019] . The majority of mice used were between 5 and 12 weeks, (supplemental table). However some strains had problems breeding, so testes from older mice were collected. We found a (small/no) age effect on MLH1 counts for male mice (Supplemental Figure)).

The only difference between preparation of spermatocyte and oocyte spreads was the volumns of the hypotonic buffer was 300$\mul instead of 3ml.

The vast majority of oocyte data was collected from neonate litters/mice between 5 to 48 hours. This approach maintain breeding pairing and still result in prophase oocytes (cite timeline of oocytes in neonates). (differences were assessed between embryo samples and neonates?). Precise staging of embryos by checking copulary plugs was difficult/not feasible in many of these wild derived strains due to their behavior. Embryonic samples were collected when pregnancy noted in females, embryos were staged based on (X table markers).

<!--
Meiocyte spreads were made following [@peters_1997]. For testes the tunica was removed and whole testes was incubated in 3ml of hypotonic solution for 45min. For ovaries after dissection, the pair of ovaries were decapsulated in cold PBS and both were incubated in 300ul of hypotonic solution for 45 min.

After incubation, gonads were transferred to sucrose solution -for mastication to make a cell slurry which was transferred to 80ul of 2% PFA solution. Cells were fixed in this solution and dried in a humid chamber at room temperature overnight. The next morning, slides were treated with a photoflow (Kodak, cite) wash.
-->

@@

The same staining protocol was applied to spermatocyte and oocyte spreads.
The staining protocol was based on X with some adaptations, and same as previously described in Peterson et al 2019.

@@

Hand measures data set generated by using ImageJ.

@@

The distributions of CO counts per cell was assessed for normailty (supplemental figure). Mouse mean MLH1 count was used to get around the bad discreet nature of count data. (normality was confirmed with X distribution, supplemental figure).

framework for analyzing the MLH1 patterns
1. write equation

 $X \sim MLH1_mouse average ~ sex * subsp + random(strain)+ strain*sex+$\xi

mixed model, observations of mouse means of mlh1, all data with both male and female observations

subspecies, sex and subsp*sex are fixed effects
strain as random effect (sample of a genetic background from subspecies)
with domesticus male set as the default

strain and sex interaction effect was included to ask / test if the effect of sex varies by strain(genetic background)

this model allows us to estimate and test the fixed effects (sex and subsp)

the random effect of strain in this model can be interpreted as the average standard deviation due to the random strain effect
and sex*strain,

From the mixed model the estimates of heterochiasmy (sexual dimorphism) are a combination of subspecies and strain specific sex effects.

We can also compare the size of the sex effects between (subspecies) and within (strain) mouse groups. (ie is the male or female strain effect larger? is there more variance due to random strain effect in male or females?)

I may have to do more research/reading on testing the random effects.

-specific statistical tests; (means, non-parametric kurts wallis t-test?) (correlation coefficient (function))
linear models lm() from X package.

-specific R packages
-non-parametric  (tests and measures for dealing with count data)
-hypothesis tested (size, DSB number)

Comparison/analysis of variation in MLH1 counts within and between groups was done within a general linear/ mix model frame work, using X package (cite).
With X formula Co count ~ blah + blah, with random strain specific sex effect.

(we used this frame work as proxy for 'polymorphism' and 'divergence' in MLH1 counts across house mouse and closely related species.)
